Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-26 ind_files/99_Appendix_B.tex
Minor rephrasing

&= \left(\frac{1}{N}\frac{\partial x_i}{\partial x_j} 
- \frac{x_i}{N^2}\sum\limits_{k=1}^{m}\frac{\partial x_k}{\partial x_j}\right)\nonumber\\
&= \frac{1}{N}\left(\delta_{ij}-p_i\right)\label{nD_jacobian_for_ito}

@@

where we have defined the total (scaled) population size\footnote{This is $N_{K}(t)$ in the main text, but we omit the subscript here to reduce notational clutter} $N = \sum_i x_i$ and the frequency of the $i\textsuperscript{th}$ type $p_i = f_i(x)$ and used the fact that $\frac{\partial x_j}{\partial x_k} = \delta_{jk}$.

@@

&= \frac{\partial}{\partial x_j}\left(\frac{\delta_{ik}}{N}-\frac{x_i}{N^2}\right)\nonumber\\
&= \frac{1}{N^2}\left(2p_i - \delta_{ij}-\delta_{ik}\right)\label{nD_hessian_for_ito}

@@

&= \frac{1}{N}\sum\limits_{j=1}^{m}\left(\delta_{ij}-p_i\right)A^{-}_{j}\nonumber\\
&= \frac{1}{N}\left(A^{-}_{i} - p_i\sum\limits_{j=1}^{m}A^{-}_{j}\right)\label{nD_for_Ito_first_term}

@@

&= \frac{1}{\sqrt{K}N^2}\sum\limits_{l=1}^{m}\left(2p_i - \delta_{ij} - \delta_{il}\right)\delta_{lk}\left(A^{+}_{l}A^{+}_{k}\right)^{\frac{1}{4}}\\
&=  \frac{1}{\sqrt{K}N^2}\left(\left(2p_i -\delta_{ij}\right)(A^{+}_{k})^{\frac{1}{2}} -\delta_{ik}\left(A^{+}_{i}A^{+}_{k}\right)^{\frac{1}{4}}\right)\\
&= \frac{1}{\sqrt{K}N^2}\left(2p_i -\delta_{ij} -\delta_{ik}\right)(A^{+}_{k})^{\frac{1}{2}}

@@

&=  \frac{1}{KN^2}\sum\limits_{j=1}^{m}\delta_{lj}\left(A^{+}_{l}A^{+}_{j}\right)^{\frac{1}{4}}(A^{+}_{k})^{\frac{1}{2}}\left(2p_i -\delta_{ij} -\delta_{ik}\right)\\
&=  \frac{1}{KN^2}(A^{+}_{k})^{\frac{1}{2}}\left(2p_i(A^{+}_{l})^{\frac{1}{2}} - (A^{+}_{i})^{\frac{1}{2}}\delta_{il} - (A^{+}_{l})^{\frac{1}{2}}\delta_{ik}\right)

@@

&= \frac{1}{KN^2}\sum\limits_{k=1}^{m}\left(2p_i(A^{+}_{k}A^{+}_{k})^{\frac{1}{2}} - (A^{+}_{i}A^{+}_{k})^{\frac{1}{2}}\delta_{ik} - (A^{+}_{k}A^{+}_{k})^{\frac{1}{2}}\delta_{ik}\right)\\
&= \frac{1}{KN^2}\left(2p_i\left(\sum\limits_{k=1}^{m} A^{+}_k\right) - 2A^{+}_{i}\right)

@@

\frac{1}{2K}\mathrm{Tr}[\mathbf{D}^{\mathrm{T}} H_{\mathbf{x}} f_i \mathbf{D}] =  \frac{-1}{KN^2}\left(A^{+}_{i}-p_i\left(\sum\limits_{k=1}^{m} A^{+}_k\right)\right)\label{nD_for_Ito_second_term}

@@

&=  \frac{1}{N\sqrt{K}}\sum\limits_{j=1}^{m}\left(\delta_{ij}-p_i\right)\left(A^{+}_{j}\right)^{1/2}dW^{(j)}_t\\
&= \frac{1}{N\sqrt{K}}\left(A^{+}_{i}\right)^{1/2}dW^{(i)}_t - p_i\sum\limits_{j=1}^{m}\left(A^{+}_{j}\right)^{1/2}dW^{(j)}_t\label{nD_for_Ito_third_term}

@@

Putting equations \eqref{nD_for_Ito_first_term}, \eqref{nD_for_Ito_second_term} and \eqref{nD_for_Ito_third_term} into \eqref{nD_Ito_formula}, we see that $p_i = f_i(\mathbf{X})_t$, the frequency of the $i\textsuperscript{th}$ type in the population $\mathbf{X}_t$, changes according to the equation:

@@

dp_i &= \underbrace{\frac{1}{N(t)}\left(A^{-}_{i} - p_i\sum\limits_{j=1}^{m}A^{-}_{j}\right)dt}_{\text{$K \to \infty$ prediction}} - \frac{1}{K}\underbrace{\frac{1}{N^2(t)}\left(A^{+}_{i}-p_i\left(\sum\limits_{k=1}^{m} A^{+}_k\right)\right)dt}_{\substack{\text{Directional finite size effects}\\\text{due to differential turnover rates}}}\\
&+ \frac{1}{\sqrt{K} N(t)}\underbrace{\left[\left(A^{+}_{i}\right)^{1/2}dW^{(i)}_t - p_i\sum\limits_{j=1}^{m}\left(A^{+}_{j}\right)^{1/2}dW^{(j)}_t\right]}_{\substack{\text{Non-directional finite size effects}\\\text{due to stochastic fluctuations}}}

@@

\frac{1}{N(t)}\left(A^{-}_{i} - p_i\sum\limits_{j=1}^{m}A^{-}_{j}\right) &=  \frac{1}{N(t)}\left[w_i(\mathbf{x})x_i + \mu Q_i(\mathbf{x})\right] - \frac{p_i}{N(t)}\sum\limits_{j=1}^{m}\left[w_j(\mathbf{x})x_j + \mu Q_j(\mathbf{x})\right]\\
&= w_i(\mathbf{x})p_i + \frac{\mu}{N(t)}Q_i(\mathbf{x}) - p_i\sum\limits_{j=1}^{m}\left[w_j(\mathbf{x})p_j + \frac{\mu}{N(t)}Q_j(\mathbf{x})\right]

@@

\frac{1}{N(t)}\left(A^{-}_{i} - p_i\sum\limits_{j=1}^{m}A^{-}_{j}\right) = (w_i(\mathbf{x}) - \overline{w})p_i + \mu\left[Q_i(\mathbf{p}) - p_i\left(\sum\limits_{j=1}^{m}Q_j(\mathbf{p})\right)\right]

@@

where we have defined $Q_j(\mathbf{p}) = Q_j(\mathbf{x})/N(t)$. Repeating the exact same calculations for the $A^{+}_i$ terms in the second term of \eqref{nD_general_stoch_freq_change} now yields equation \eqref{nD_eqn_for_frequencies} in the main text.

>>>>

&= \left(\left(\frac{1}{\sum\limits_{r=1}^{m}x_r}\right)\frac{\partial x_i}{\partial x_j} 
- \left(\frac{x_i}{\left(\sum\limits_{r=1}^{m}x_r\right)^2}\right)\sum\limits_{k=1}^{m}\frac{\partial x_k}{\partial x_j}\right)\nonumber\\
&= \frac{1}{\sum\limits_{r=1}^{m}x_r}\left(\delta_{ij}-p_i\right)\label{nD_jacobian_for_ito}

@@

where we have defined the frequency of the $i\textsuperscript{th}$ type $p_i = f_i(\mathbf{x})$ and used the fact that $\frac{\partial x_j}{\partial x_k} = \delta_{jk}$.

@@

&= \frac{\partial}{\partial x_j}\left(\frac{\delta_{ik}}{\sum\limits_{r=1}^{m}x_r}-\frac{x_i}{\left(\sum\limits_{r=1}^{m}x_r\right)^2}\right)\nonumber\\
&= \frac{1}{\left(\sum\limits_{r=1}^{m}x_r\right)^2}\left(2p_i - \delta_{ij}-\delta_{ik}\right)\label{nD_hessian_for_ito}

@@

&= \frac{1}{\sum\limits_{r=1}^{m}x_r}\sum\limits_{j=1}^{m}\left(\delta_{ij}-p_i\right)A^{-}_{j}\nonumber\\
&= \frac{1}{\sum\limits_{r=1}^{m}x_r}\left(A^{-}_{i} - p_i\sum\limits_{j=1}^{m}A^{-}_{j}\right)\label{nD_for_Ito_first_term}

@@

&= \frac{1}{\sqrt{K}\left(\sum\limits_{r=1}^{m}x_r\right)^2}\sum\limits_{l=1}^{m}\left(2p_i - \delta_{ij} - \delta_{il}\right)\delta_{lk}\left(A^{+}_{l}A^{+}_{k}\right)^{\frac{1}{4}}\\
&=  \frac{1}{\sqrt{K}\left(\sum\limits_{r=1}^{m}x_r\right)^2}\left(\left(2p_i -\delta_{ij}\right)(A^{+}_{k})^{\frac{1}{2}} -\delta_{ik}\left(A^{+}_{i}A^{+}_{k}\right)^{\frac{1}{4}}\right)\\
&= \frac{1}{\sqrt{K}\left(\sum\limits_{r=1}^{m}x_r\right)^2}\left(2p_i -\delta_{ij} -\delta_{ik}\right)(A^{+}_{k})^{\frac{1}{2}}

@@

&=  \frac{1}{K\left(\sum\limits_{r=1}^{m}x_r\right)^2}\sum\limits_{j=1}^{m}\delta_{lj}\left(A^{+}_{l}A^{+}_{j}\right)^{\frac{1}{4}}(A^{+}_{k})^{\frac{1}{2}}\left(2p_i -\delta_{ij} -\delta_{ik}\right)\\
&=  \frac{1}{K\left(\sum\limits_{r=1}^{m}x_r\right)^2}(A^{+}_{k})^{\frac{1}{2}}\left(2p_i(A^{+}_{l})^{\frac{1}{2}} - (A^{+}_{i})^{\frac{1}{2}}\delta_{il} - (A^{+}_{l})^{\frac{1}{2}}\delta_{ik}\right)

@@

&= \frac{1}{K\left(\sum\limits_{r=1}^{m}x_r\right)^2}\sum\limits_{k=1}^{m}\left(2p_i(A^{+}_{k}A^{+}_{k})^{\frac{1}{2}} - (A^{+}_{i}A^{+}_{k})^{\frac{1}{2}}\delta_{ik} - (A^{+}_{k}A^{+}_{k})^{\frac{1}{2}}\delta_{ik}\right)\\
&= \frac{1}{K\left(\sum\limits_{r=1}^{m}x_r\right)^2}\left(2p_i\left(\sum\limits_{k=1}^{m} A^{+}_k\right) - 2A^{+}_{i}\right)

@@

\frac{1}{2K}\mathrm{Tr}[\mathbf{D}^{\mathrm{T}} H_{\mathbf{x}} f_i \mathbf{D}] =  \frac{-1}{K\left(\sum\limits_{r=1}^{m}x_r\right)^2}\left(A^{+}_{i}-p_i\left(\sum\limits_{k=1}^{m} A^{+}_k\right)\right)\label{nD_for_Ito_second_term}

@@

&=  \frac{1}{\left(\sum\limits_{r=1}^{m}x_r\right)\sqrt{K}}\sum\limits_{j=1}^{m}\left(\delta_{ij}-p_i\right)\left(A^{+}_{j}\right)^{1/2}dW^{(j)}_t\\
&= \frac{1}{\left(\sum\limits_{r=1}^{m}x_r\right)\sqrt{K}}\left(A^{+}_{i}\right)^{1/2}dW^{(i)}_t - p_i\sum\limits_{j=1}^{m}\left(A^{+}_{j}\right)^{1/2}dW^{(j)}_t\label{nD_for_Ito_third_term}

@@

Putting equations \eqref{nD_for_Ito_first_term}, \eqref{nD_for_Ito_second_term} and \eqref{nD_for_Ito_third_term} into \eqref{nD_Ito_formula} and letting $N_K(t) = \sum\limits_{r=1}^{m}x_r$ we see that $p_i = f_i(\mathbf{X})_t$, the frequency of the $i\textsuperscript{th}$ type in the population $\mathbf{X}_t$, changes according to the equation:

@@

dp_i &= \underbrace{\frac{1}{N_K(t)}\left(A^{-}_{i} - p_i\sum\limits_{j=1}^{m}A^{-}_{j}\right)dt}_{\text{$K \to \infty$ prediction}} - \frac{1}{K}\underbrace{\frac{1}{N_K^2(t)}\left(A^{+}_{i}-p_i\left(\sum\limits_{k=1}^{m} A^{+}_k\right)\right)dt}_{\substack{\text{Directional finite size effects}\\\text{due to differential turnover rates}}}\\
&+ \frac{1}{\sqrt{K} N_K(t)}\underbrace{\left[\left(A^{+}_{i}\right)^{1/2}dW^{(i)}_t - p_i\sum\limits_{j=1}^{m}\left(A^{+}_{j}\right)^{1/2}dW^{(j)}_t\right]}_{\substack{\text{Non-directional finite size effects}\\\text{due to stochastic fluctuations}}}

@@

\frac{1}{N_K(t)}\left(A^{-}_{i} - p_i\sum\limits_{j=1}^{m}A^{-}_{j}\right) &=  \frac{1}{N_K(t)}\left[w_i(\mathbf{x})x_i + \mu Q_i(\mathbf{x})\right] - \frac{p_i}{N_K(t)}\sum\limits_{j=1}^{m}\left[w_j(\mathbf{x})x_j + \mu Q_j(\mathbf{x})\right]\\
&= w_i(\mathbf{x})p_i + \frac{\mu}{N_K(t)}Q_i(\mathbf{x}) - p_i\sum\limits_{j=1}^{m}\left[w_j(\mathbf{x})p_j + \frac{\mu}{N_K(t)}Q_j(\mathbf{x})\right]

@@

\frac{1}{N_K(t)}\left(A^{-}_{i} - p_i\sum\limits_{j=1}^{m}A^{-}_{j}\right) = (w_i(\mathbf{x}) - \overline{w})p_i + \mu\left[Q_i(\mathbf{p}) - p_i\left(\sum\limits_{j=1}^{m}Q_j(\mathbf{p})\right)\right]

@@

where we have defined $Q_j(\mathbf{p}) = Q_j(\mathbf{x})/N_K(t)$. Repeating the exact same calculations for the $A^{+}_i$ terms in the second term of \eqref{nD_general_stoch_freq_change} now yields equation \eqref{nD_eqn_for_frequencies} (which is also equation \eqref{nD_stochastic_RM}) in the main text.

### 2023-03-26 ind_files/99_Appendix_D_examples.tex
Minor rephrasing

\section{An example in one dimension: The stochastic logistic equation}
Here, we analyze example \ref{ex_1D_stoch_logistic}. To recap, we had a population of individuals that exhibit a constant per-capita birth rate $\lambda > 0$, and a per-capita death rate that had the linear density-dependence $d(n) = \left(\mu + (\lambda-\mu)\frac{n}{K}\right)n$, where $\mu$ and $K$ are positive constants. Thus, we have the equations
\begin{equation}
\label{ex_1D_stoch_logistic_BD_eqns}
\begin{aligned}
    b(n) &= \lambda n\\
    d(n) &= \left(\mu + (\lambda-\mu)\frac{n}{K}\right)n
\end{aligned}
\end{equation}
Here, $K$ is the system-size parameter. Introducing the new variable $x=n/K$, we obtain
\begin{align*}
    b_K(x) &= \frac{1}{K}b(n) = \frac{1}{K}\lambda Kx\\
    d_K(x) &= \frac{1}{K}d(n) = \frac{1}{K}\left(\mu + (\lambda-\mu)\frac{Kx}{K}\right)Kx
\end{align*}
Thus, we have
\begin{equation*}
    A^{\pm}(x) = b_K(x)\pm d_K(x) = x\left(\lambda \pm \left(\left(\mu + (\lambda-\mu)x\right)\right) \right)
\end{equation*}
Defining $r=\lambda-\mu$ and $v=\lambda+\mu$ and using equation \eqref{1D_SDE}, we 
see that the `mesoscopic view' of the system is given by the solution of the SDE
\begin{equation}\label{ex_1D_stoch_logistic_full_SDE}
dX_t =  rX_t(1-X_t)dt + \sqrt{\frac{X_t(v+rX_t)}{K}}dW_t
\end{equation}
From equation \eqref{1D_det_limit}, we see that the deterministic dynamics are
\begin{equation}\label{ex_1D_stoch_logistic_det_limit}
\frac{dx}{dt} = A^-(x) = rx(1-x)
\end{equation}
showing that in the infinite population limit, we obtain the logistic equation. This derivation also makes it clear that two systems with very different stochastic dynamics can nevertheless converge to the same infinite population limit \eqref{ex_1D_stoch_logistic_det_limit}, since \eqref{ex_1D_stoch_logistic_det_limit}t only depends on the difference $\lambda-\mu$.
\myfig{0.6}{figures/App_stoch_logistic_same_w_different_t.png}{\textbf{Comparison of time series produced by Gillespie simulations of the stochastic logistic equation for two different parameter values.} Simulations with the parameter values $\lambda = 4, \mu = 2$ are plotted in red, and simulations with the parameter values $\lambda = 31, \mu = 29$ are plotted in blue. Both simulations have $K=200$. The infinite population limit of both simulations is plotted in black dotted lines. The graph shows 10 realizations each for the two sets of parameter values.}{fig_1D_same_w_different_t_comparison}
To illustrate the effects of this seemingly innocent fact, figure \ref{fig_1D_same_w_different_t_comparison} compares two simulations which have the same value of $\lambda - \mu$ but a ten-fold difference in $\lambda+\mu$. As is clear from the figure, though both populations have the same behavior at the infinite population limit, populations with a higher value of $\lambda+\mu$ exhibit much wilder fluctuations and are therefore more prone to stochastic extinction. This is the root cause of the noise-induced selection that occurs in higher dimensions, discussed in detail in part \ref{part_summary}.

>>>>

\section{An example in one dimension: The stochastic logistic equation}
Here, we analyze example \ref{ex_1D_stoch_logistic}. To recap, we had a population of individuals that exhibit a constant per-capita birth rate $\lambda > 0$, and a per-capita death rate that had the linear density-dependence $d(n) = \left(\mu + (\lambda-\mu)\frac{n}{K}\right)n$, where $\mu$ and $K$ are positive constants. Thus, we have the equations
\begin{equation}
\label{ex_1D_stoch_logistic_BD_eqns}
\begin{aligned}
    b(n) &= \lambda n\\
    d(n) &= \left(\mu + (\lambda-\mu)\frac{n}{K}\right)n
\end{aligned}
\end{equation}
Here, $K$ is the system-size parameter. Introducing the new variable $x=n/K$, we obtain
\begin{align*}
    b_K(x) &= \frac{1}{K}b(n) = \frac{1}{K}\lambda Kx\\
    d_K(x) &= \frac{1}{K}d(n) = \frac{1}{K}\left(\mu + (\lambda-\mu)\frac{Kx}{K}\right)Kx
\end{align*}
Thus, we have
\begin{equation*}
    A^{\pm}(x) = b_K(x)\pm d_K(x) = x\left(\lambda \pm \left(\left(\mu + (\lambda-\mu)x\right)\right) \right)
\end{equation*}
Defining $r=\lambda-\mu$ and $v=\lambda+\mu$ and using equation \eqref{1D_SDE}, we 
see that the `mesoscopic view' of the system is given by the solution of the SDE
\begin{equation}\label{ex_1D_stoch_logistic_full_SDE}
dX_t =  rX_t(1-X_t)dt + \sqrt{\frac{X_t(v+rX_t)}{K}}dW_t
\end{equation}
From equation \eqref{1D_det_limit}, we see that the deterministic dynamics are
\begin{equation}\label{ex_1D_stoch_logistic_det_limit}
\frac{dx}{dt} = A^-(x) = rx(1-x)
\end{equation}
showing that in the infinite population limit, we obtain the logistic equation. This derivation also makes it clear that two systems with very different stochastic dynamics can nevertheless converge to the same infinite population limit \eqref{ex_1D_stoch_logistic_det_limit}, since equation \eqref{ex_1D_stoch_logistic_det_limit} only depends on the difference $\lambda-\mu$.
\myfig{0.6}{figures/App_stoch_logistic_same_w_different_t.png}{\textbf{Comparison of time series produced by Gillespie simulations of the stochastic logistic equation for two different parameter values.} Simulations with the parameter values $\lambda = 4, \mu = 2$ are plotted in red, and simulations with the parameter values $\lambda = 31, \mu = 29$ are plotted in blue. Both simulations have $K=200$. The infinite population limit of both simulations is plotted in black dotted lines. The graph shows 10 realizations each for the two sets of parameter values.}{fig_1D_same_w_different_t_comparison}
To illustrate the effects of this seemingly innocent fact, figure \ref{fig_1D_same_w_different_t_comparison} compares two simulations which have the same value of $\lambda - \mu$ but a ten-fold difference in $\lambda+\mu$. As is clear from the figure, though both populations have the same behavior at the infinite population limit, populations with a higher value of $\lambda+\mu$ exhibit much wilder fluctuations and are therefore more prone to stochastic extinction. This is the root cause of the noise-induced selection that occurs in higher dimensions, discussed in detail in part \ref{part_summary}.
